|

### 2024-01-22 rmds/manuscript.Rmd
updates

---
title: |
  Quantifying maternal reproductive output of chondrichthyan fishes
subtitle: |
author:
- name: Alastair V Harry
  affiliation: Fisheries & Agriculture Resource Management, Department of Primary Industries and Regional Development, 39 Northside Drive, Hillarys, 6025, WA, Australia
  email: alastair.harry@gmail.com
  number: 1
- name: Ivy Baremore
  affiliation: MarAlliance
  number: 2
- name: Andrew Piercy
  affiliation: Valencia College
  number: 3
abstract: |
  For the live-bearing and egg-laying class of chondrichthyan fishes a three parameter logistic 'maternity' function with a variable upper asymptote, $P_{Max}$, can be used to calculate the average probability of a female giving birth or laying eggs in a season. Although fundamental to calculating the reproductive capacity of a population, relatively few studies report maternity functions. Instead, maturity functions have typically been used as a proxy, despite evidence of a delay between attainment of functional maturity and pregnancy in some species. This study examined the relative performance of alternative approaches for quantifying maternal reproductive output. Applying logistic models to a combination of simulated and empirical data showed that it was feasible to estimate $P_{Max}$ from data and that precision, bias, and confidence interval coverage often improved compared to when a fixed value was used. At sample sizes < 250 individuals a fixed-value for $P_{Max}$ was more effective at reducing bias. Maturity parameters could be estimated with greater precision however substituting them in place of maternity parameters overestimated lifetime reproductive output. The extent to which maturity functions can provide a good approximation for maternity functions may be species specific and requires futher research. Greater use of maternity functions has the potential to improve calculation of reproductive output in quantitative populations models. We proposed that estimation of $P_{Max}$ is generally preferable to using a fixed value. In addition to improvements in parameter estimation, this method involves fewer assumptions and enables statistical inferences to be made on frequency of reproduction.

@@

In chondrichthyan fishes (sharks, rays, skates, and chimaeras), the process of maturation can likewise be modeled with a simple 2PLF. However for the intents and purposes of calculating reproductive capacity, chondrichthyans differ in several ways to most teleosts. Firstly, in many chondrichthyans the duration of the reproductive cycle is longer than a year meaning the proportion of females that will give birth in the population in a given season is less than one [@frisk_life_2005; @dulvy_you_2008; @rigby_patterns_2015]. Secondly, there may be a lag between when females reach sexual maturity and when they begin reproducing [@harry_age_2013; @fujinami_reproductive_2017]. This is confounded by the adopted definition of maturity, which can lead to variable estimates of size and age at maturity [@braccini_determining_2006; @walker_spatial_2007; @montealegre-quijano_sexual_2014]. Finally, because female fecundity is low (usually $\ll$ 100 embryos), total pup or egg production and recruitment are likely to be closely related to the number of reproductively active females, exhibiting little interannual variability [@taylor_stock-recruitment_2013]. This contrasts the often highly variable relationship between stock size and recruitment in teleost fish, and warrants a precise definition and calculation of reproductive output. For these reasons, unlike in teleosts, a maturity function may not be the most appropriate tool to quantify reproductive capacity.

@@

Numerous authors have adopted Walker's [-@walker_reproduction_2005] maternity function, including for species with biennial or longer reproductive cycles [@huveneers_reproductive_2007; @rochowski_reproductive_2015; @trinnie_reproductive_2016] as well as those with annual or shorter reproductive cycles [@trinnie_reproductive_2009; @mejia-falla_reproductive_2012; @harry_age_2013; @taylor_living_2016]. Techniques have been described for determining maternal output in asynchronously reproducing species [@braccini_determining_2006; @colonello_reproductive_2016], and maternity functions are also increasingly being used directly in shark and ray population assessments [@sedar_sedar_2012; @sedar_sedar_2017].

Despite an increase in use, only a small fraction of reproductive studies in recent years have reported maternity functions, potentially due to a range of factors. Classification of maternal condition is more data-intensive than maturity condition, ideally requiring monthly sampling over a year or longer in order to establish the timing and duration of the female ovarian and uterine cycles. Such data can be difficult to collect for sharks and rays, which are often data-poor and sampled opportunistically. The teleost-oriented foundations of fisheries science have also contributed to the general lack of awareness of maternity functions; Walker (2005) is the sole description on this type of analysis for chondrichthyans and there are no primary literature sources that describe specific methods for chondrichthyan fishes.

@@

To assess the feasibility of estimating maternity parameters a simulation study was done using the 3PLF to generate data for populations with varying reproductive frequencies (Figure 1). Simulations were conducted for a range of gear selectivities and sample sizes reflective of those typically available in reproductive studies and fishery sampling programs. The performance of three methods in estimating 'true' maternity parameters from simulated data was tested by examining precision, bias, and interval coverage. To illustrate the effects of misspecifying maternity parameters, lifetime reproductive output was calculated for each simulated population using parameters derived from each of the methods, and by substituting maturity parameters for maternity parameters. Finally, empirical data from previously published studies were reanalysed to illustrate possible strategies for estimating maternity parameters and making inferences on reproductive frequency. All simulation modelling and data analysis was conducted using the R language [@rcoreteam_r:_2018] and the Template Model Builder (TMB) R package [@kristensen_tmb:_2016].

@@

Catch at length data were simulated using a female-only, length-based catch curve model [@L3Assess]. The approach involved calculating expected survival and catches per recruit subject to specified life history parameters, gear selectivity, and total mortality. Growth was modelled using a von Bertalanffy growth function with length *l*, of a fish of age, *a*, calculated as $$l(a) = L_\infty (1- e^{-K(a-a_0)})$$

where $L_\infty$ is asymptotic length, *K* is the growth coefficient and $a_0$ is the hypothetical age at zero length. Recruitment into the population was assumed to occur at age 0 with length conforming to a normal distribution. The proportion of fish in length class, *l*, at age, $a = 0$, was calculated as $$\theta_k = \int_{l^-_k}^{l^+_k}f_{a=0}(l)dl$$ where $l^-$ and $l^+$ are the lower and upper limits of each 1cm length class, *k*. $f_{a=0}(l)$ is the value of the normal probability density function at age $a = 0$ for a given length, *l*,$$f_{a=0}(l) = \frac{1}{\sigma\sqrt{2{\pi}}}\exp\bigg[\frac{(l-\mu)^2}{2\sigma^2}\bigg]
$$

@@

Growth in subsequent age classes was modelled using a length transition matrix [@punt1997; @hall2000], $G = g_{j,k}$ that represents the probability that a fish in length class, $j$, will grow into length class, $k$, over a specified time interval $$ G = \begin{bmatrix}
 g_{1,1}&0& 0 &0\\ 
 g_{2,1}& g_{2,2} & 0  &0 \\ 
 \cdots&\cdots  & \ddots  & 0\\ 
 g_{n,1}& g_{n,2} & \cdots & g_{n,n}
\end{bmatrix}$$

where the elements of the matrix followed the general form of Punt *et al* [-@punt1997] and are given by$$ g_{k,j} = \left\{
    \begin{array}{ll}
        \int_{{\infty}^-}^{\mathrm{l_{k}^+}}f[\phi(l,j)]dl & \mbox{if }k=1 \\
        \int_{l_{k}^-}^{\mathrm{l_{k}^+}}f[\phi(l,j)]dl &\mbox{if }k=1<k<n \\
\int_{l_{k}^-}^{\mathrm{\infty^+}}f[\phi(l,j)]dl &\mbox{if }k=n \\
    \end{array}
\right.
$$

@@

Selectivity in the model was assumed to follow a gamma function based on gillnet selectivity experiments [@kirkwood_gill_1986; @punt_stock_1998], where the selectivity of an individual in length class, $l$ was given by $$S(l) = \left [ \frac{l}{\theta_1 \theta_2} \right ]^{\theta_1 }exp(\theta_1-\frac{l}{\theta_2}) $$ where $\theta_1$ and $\theta_2$ are estimated parameters [@kirkwood_gill_1986]. Two selectivity scenarios were simulated corresponding to low and high selectivity of the maternal component of the population. For consistency between species, gillnet mesh sizes were chosen that would result in peak relative selectivity occurring at the lengths equal to 5% and 95% of the population in maternal condition (Table 1).

@@

$$ N_{l,a} = \left\{    \begin{array}{ll}        1 & \mbox{if }a=0 \\        N_{l,a-1}\exp(-Z_{l}) &\mbox{if }a=1<a<A \\        N_{l,a-1}\exp(-Z_{l})/(1-\exp(-Z_{l})) &\mbox{if }a=A \\    \end{array}\right.$$

The estimated numbers of fish caught in length class, l, at age, a, was calculated using the Baranov catch equation as

@@

Next, maturity-at-length data, $\psi(l)$, were randomly generated by simulating from a Bernoulli distribution, $\Psi(l) \sim B(P)$, where $P$ was given by the 2PLF: $$P =  \left ( 1 + e^{-ln(19)\left ( \frac{l - L_{50}}{L_{95} - L_{50}} \right )}  \right )^{-1}$$ This processes was then repeated to generate maternity-at-length data, $\Psi'(l)$, using the 3PLF: $$\Psi' \sim B(P)$$ $$P =  P_{Max}\left ( 1 + e^{-ln(19)\left ( \frac{l(a) - L'_{50}}{L'_{95} - L'_{50}} \right )}  \right )^{-1}$$ Finally, 250 iterations were run for each of the 144 unique variables combinations.

@@

Two approaches to estimating maternity parameters were compared, the 3PLF itself (3PLF-free) and the 3PLF function with a fixed asymptote (3PLF-fixed). Additionally, two approaches using a 2PLF were also examined, using maternity-at-length data (2PLF-maternity) and maturity-at-length data (2PLF maturity). The 3PLF-free method was examined to validate the utility of this model, which has so far been used in only a single study [@colonello_reproductive_2016]. The 3PLF-fixed is the method described by Walker [-@walker_reproduction_2005] and most commonly used in practice. The 2PLF-maternity method was used to examine what effect simply ignoring the upper asmyptote had (i.e. $P_{Max}$ fixed at 1). The 2PLF-maturity method is tacitly assumed if using mature females to approximate maternal ones. The resulting maturity curve can then by weighted by the proportion of gravid females to calculate annual reproductive output (although in practice annual fecundity is more often modified, for example halved for a species suspected to reproduce biennially).

Only the 3PLF-free method involves statistically estimating $P_{Max}$, which must be subjectively chosen in the case of the 3PLF-fixed and the 2PLF-maturity. $P_{Max}$ is ideally chosen based on detailed study of the ovarian and uterine cycles [@walker_reproduction_2005], or alternatively based on the proportion of mature females observed to be in maternal condition during sampling [@baremore_reproduction_2012; @harry_age_2013; @trinnie_reproductive_2016]. For the purposes of the simulation, the timing and duration of the uterine and ovarian cycles were assumed to be unknown. The following procedure was used to 'guess' the fixed value of $P_{Max}$ in each simulation. $P_{Max}$ was chosen as the proportion of females in maternal condition above a the length at which $99\%$ of females were mature, $L > \Psi(L_{99\%})$. If this failed (e.g. zero pregnant females), $L_{95}$ was used, followed by $L_{50}$, and finally the proportion of all mature females in the sample that were in maternal condition. Statistical estimation of all non-fixed parameters in each of the methods was undertaken using maximum likelihood. Nonparametric bootstrapping was used to calculate approximate 50% confidence intervals for the best-fit parameters from 250 resampled data sets.

@@

An empirical case study was also undertaken to show how maternity functions can be used in practice. Data from two studies of the reproductive biology of the sandbar shark, *Carcharhinus plumbeus*, in the Gulf of Mexico and western North Atlantic Ocean were combined and reanalysed using both 3PLF methods. The sandbar shark is a relatively long-lived ($A_{Max}$ = 33 years) and slow growing species ($K = 0.12 \text{yr}^{-1}$) that was thought to have a biennial reproductive cycle [@springer_natural_1960]. The Gulf of Mexico and western North Atlantic Ocean population has been considered overfished since the late 1970s [@sminkey_age_1995] and during the 2000s detailed studies of sandbar shark reproductive biology and growth were undertaken for stock assessment purposes [@baremore_reproduction_2012; @piercy_histological_2016]. In their study, Baremore and Hale [-@baremore_reproduction_2012] estimated maternity parameters using the 2PLF-maternity method, weighting the final curve by 0.37 based on the proportion of pregnant females observed, most closely aligning with a triennial reproductive cycle. Piercy *et al* [-@piercy_histological_2016] did not estimate maternity parameters but also suggested that the average reproductive cycle was likely to be longer than two years based on examination of ovarian follicles. To examine relative support for either a biennial or triennial reproductive cycle, 3PLF-fixed models were fit with $P_{Max}$ estimated and fixed at 0.5 and 0.33, respectively.

@@

Parameter estimation was generally straightforward; with sample sizes $> 100$ a success rate of $\geq 95\%$ was achieved in all scenarios except the 2PLF-maternity method for gummy shark (Figure S1, Table S1). Overall, convergence and parameter boundary failures were most common with the 2PLF-maternity method which essentially involved fitting data to an under-parameterized model where $P_{Max}$ was always fixed at 1. In addition to convergence failures, 7 simulations also generated insufficient maternal data (zero or one maternal females in the simulated data), and were discarded.

@@

For the 3PLF methods, parameter estimates tended to be biased high at lower sample sizes (Figure S2, Figure S3). Fixing the asymptote was effective at reducing bias at sample sizes $\leq$ 250, although above this neither method was clearly preferable. The accuracy of parameter estimates varied considerably across different combinations of variables used in the simulation and among parameters (Figure S4, Figure S5). For $L'_{50}$ gains in precision tended to diminish at sample sizes $\geq$ 500. $P_{Max}$ was comparatively more challenging to estimate accurately, with precision continuing to improve noticeably at the largest sample sizes (Figure 3).

@@

data %>% 
  filter(par_name %in% "c") %>% 
  filter(method_long %in% c("3PLF - free", "3PLF - fixed")) %>% 
  filter(mesh_name %in% "high") %>% 
  ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(as.factor(Nsamples), abs_error, col = repro_freq, group = repro_freq)) +
  geom_line(aes(as.factor(Nsamples), abs_error, col = repro_freq, group = repro_freq)) + 
  facet_grid(species ~ method_long) + 
  scale_color_brewer(palette = "Spectral") + 
  guides(colour = guide_legend(bquote(atop(italic(P)[Max],("Reproductive frequency"))))) + 
  theme_bw() + 
  labs(x = "Sample size", y = "Mean % absolute relative error") +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = "center")
```

While the 3PLF-fixed method was in some cases able to reduce bias at lower sample sizes, a trade off in the use of this method was poorer interval coverage (Figure 4, Figure S6). For most simulations, interval coverage for $L'_{50}$ was well below the expected level of 50%, irrespective of sample size. In contrast, for the 3PLF-free method interval coverage oscillated around 50% for both $L'_{50}$ and $P_{Max}$.

```{r fig.width = 7, fig.height= 5.5, fig.cap = "Interval coverage for $\\hat{L'_{50}}$ for 3PLF methods. Figure shows the percentage of simulations where the true parameter value fell within the 50% bootstrap confidence interval. "}
data %>% 
  filter(par_name %in% "m50", method_long %in% c("3PLF - free", "3PLF - fixed")) %>% 
  filter(mesh_name %in% "high") %>% 
  ggplot() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_point(aes(as.factor(Nsamples), int_coverage, col = repro_freq, group = repro_freq)) +
  geom_line(aes(as.factor(Nsamples), int_coverage, col = repro_freq, group = repro_freq)) + 
  facet_grid(species ~ method_long, scales = "free_y") + 
  scale_color_brewer(palette = "Spectral") + 
  guides(colour = guide_legend(bquote(atop(italic(P)[Max],("Reproductive frequency"))))) + 
  theme_bw() + 
  labs(x = "Sample size", y = "% interval coverage") +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = "center")
```

Differences in life history and gear selectivity played an important role in the ability to obtain precise and unbiased maternity parameters. Certain combinations of variables in the simulated data resulted in low numbers of females in maternal or immature condition making parameter estimation difficult (Figure S8, Figure S9). These effects were not necessarily consistent between species and seemed to reflect differences in the underlying population length structure. For example, in high selectivity scenarios for gummy sharks \~10% of individuals were immature (Figure S9) compared to \~50% in the corresponding scenarios for school sharks. As $P_{Max}$ decreased the proportion of both maternal females and immature females in the analysis also decreased. The latter effect was due to the right shift in length structure associated with the proportionally fewer juveniles in the population.

Overall, the 2PLF-maternity method performed the worst. Ignoring $P_{Max}$ was equivalent to introducing a fixed bias in this parameter that increased in magnitude as reproductive periodicity increased (Figure S3). This manifested in an overestimation of $L'_{50}$ that was exacerbated by gear selectivity effects (Figure S2). In contrast, using a maturity function to approximate the maternity function (2PLF-maturity method) resulted in relatively good performance. $L_{50}$ could usually be estimated with a higher precision than $L'_{50}$. For the gummy shark, where $L_{50}$ and $L'_{50}$ were relatively similar, using the 2PLF-maturity method led to a constant underestimate of $L'_{50}$ of approximately 2% and exceeding the performance of the 3PLF-methods at sample sizes \< 250.

The effect of the different methods in ultimately calculating $R_0$ varied considerably and was strongly influenced by life history and selectivity characteristics. Greater precision was achieved with the 3PLF methods, with the 3PLF-free method generally performing best (Figure 5, Figure S11). For gummy sharks the 3PLF-fixed method was the preferred model at lower sample sizes and lower values of $P_{Max}$, consistent with decreasing performance of 3PLF methods as $P_{Max}$ decreased (Figure 2, Figure 3).

>>>>

---
title: |
  Quantifying maternal reproductive output of chondrichthyan fishes
subtitle: |
author:
- name: Alastair V Harry
  affiliation: Fisheries & Agriculture Resource Management, Department of Primary Industries and Regional Development, 39 Northside Drive, Hillarys, 6025, WA, Australia
  email: alastair.harry@gmail.com
  number: 1
- name: Ivy Baremore
  affiliation: MarAlliance, Roatán, Islas de la Bahía, Honduras
  number: 2
- name: Andrew Piercy
  affiliation: Valencia College
  number: 3
abstract: |
  For the live-bearing and egg-laying class of chondrichthyan fishes a three parameter logistic 'maternity' function with a variable upper asymptote, $P_{Max}$, can be used to calculate the average probability of a female giving birth or laying eggs in a season. Although fundamental to calculating the reproductive capacity of a population, relatively few studies report maternity functions. Instead, maturity functions have typically been used as a proxy, despite evidence of a delay between attainment of functional maturity and pregnancy in some species. This study examined the relative performance of alternative approaches for quantifying maternal reproductive output. Applying logistic models to a combination of simulated and empirical data showed that it was feasible to estimate $P_{Max}$ from data and that precision, bias, and confidence interval coverage often improved compared to when a fixed value was used. At sample sizes < 250 individuals a fixed-value for $P_{Max}$ was more effective at reducing bias in simulated data for gummy sharks, *Mustelus antarcticus*. Maturity parameters could be estimated with greater precision however substituting them in place of maternity parameters overestimated lifetime reproductive output. The extent to which maturity functions can provide a good approximation for maternity functions may be species specific and requires futher research. Greater use of maternity functions has the potential to improve calculation of reproductive output in quantitative populations models. We proposed that estimation of $P_{Max}$ is generally preferable to using a fixed value. In addition to improvements in parameter estimation, this method involves fewer assumptions and enables statistical inferences to be made on frequency of reproduction.

@@

In chondrichthyan fishes (sharks, rays, skates, and chimaeras), the process of maturation can be modelled in a similar manner to teleost fishes with a simple 2PLF; however, for the intents and purposes of calculating reproductive capacity, chondrichthyans differ in several ways to most teleosts. Firstly, in many chondrichthyans the duration of the reproductive cycle is longer than a year, meaning the proportion of females that will give birth in the population in a given season is less than one [@frisk_life_2005; @dulvy_you_2008; @rigby_patterns_2015]. Secondly, there may be a lag between when females reach sexual maturity and when they begin reproducing [@harry_age_2013; @fujinami_reproductive_2017]. This is confounded by the adopted definition of maturity, which can lead to variable estimates of size and age at maturity [@braccini_determining_2006; @walker_spatial_2007; @montealegre-quijano_sexual_2014]. Finally, because female fecundity is low (usually $\ll$ 100 embryos), total pup or egg production and recruitment are likely to be closely related to the number of reproductively active females, exhibiting little interannual variability [@taylor_stockrecruitment_2013]. This contrasts the often highly variable relationship between stock size and recruitment in teleost fish, and warrants a precise definition and calculation of reproductive output. For these reasons, unlike in teleosts, a maturity function may not be the most appropriate tool to quantify reproductive capacity.

@@

Numerous authors have adopted Walker's [-@walker_reproduction_2005] maternity function, including for species with biennial or longer reproductive cycles [@huveneers_reproductive_2007; @rochowski_reproductive_2015; @trinnie_reproductive_2016] as well as those with annual or shorter reproductive cycles [@trinnie_reproductive_2009; @mejiafalla_reproductive_2012; @harry_age_2013; @taylor_living_2016]. Techniques have been described for determining maternal output in asynchronously reproducing species [@braccini_determining_2006; @colonello_reproductive_2016], and maternity functions are also increasingly being used directly in shark and ray population assessments [@sedar_sedar_2012; @sedar_sedar_2017; @sedar_sedar_2023].

Despite an increase in use, only a small fraction of reproductive studies in recent years have reported maternity functions. Classification of maternal condition is more data-intensive than maturity condition, ideally requiring monthly sampling over a year or longer in order to establish the timing and duration of the female ovarian and uterine cycles. Such data can be difficult and costly to collect for sharks and rays, which are often data-poor and sampled opportunistically. The teleost-oriented foundations of fisheries science have also contributed to the general lack of awareness of maternity functions; Walker (2005) is the sole description on this type of analysis for chondrichthyans and there are no primary literature sources that describe specific methods for chondrichthyan fishes.

@@

To assess the feasibility of estimating maternity parameters a simulation study was carried out using the 3PLF to generate data for populations with varying reproductive frequencies (Figure 1). Simulations were conducted for a range of gear selectivities and sample sizes reflective of those typically available in reproductive studies and fishery sampling programs. The performance of three methods in estimating 'true' maternity parameters from simulated data was tested by examining precision, bias, and interval coverage. To illustrate the effects of misspecifying maternity parameters, lifetime reproductive output was calculated for each simulated population using parameters derived from each of the methods, and by substituting maturity parameters for maternity parameters. Finally, empirical data from previously published studies were reanalysed to illustrate possible strategies for estimating maternity parameters and making inferences on reproductive frequency. All simulation modelling and data analysis was conducted using the R language [@r_core_team_r:_2018] and the Template Model Builder (TMB) R package [@kristensen_tmb:_2016].

@@

Catch at length data were simulated using a female-only, length- and age-based equilibrium population model [@hesp_l3assess_2023]. The approach involved calculating expected survival and catches per recruit subject to specified life history parameters, gear selectivity, and total mortality. Growth was modelled using a von Bertalanffy growth function with length *l*, of a fish of age, *a*, calculated as $$l(a) = L_\infty (1- e^{-K(a-a_0)})$$

where $L_\infty$ is asymptotic length, *K* is the growth coefficient and $a_0$ is the hypothetical age at zero length. Recruitment into the population was assumed to occur at age 0 with length conforming to a normal distribution. The proportion of fish of length, *l*, at age, $a = 0$, was calculated as $$\Theta_k = \int_{l^-_k}^{l^+_k}f_{a=0}(l)dl$$ where $l^-$ and $l^+$ are the lower and upper limits of each 1cm length class, *k*. $f_{a=0}(l)$ is the value of the normal probability density function at age $a = 0$ for a given length, *l*,$$f_{a=0}(l) = \frac{1}{\sigma\sqrt{2{\pi}}}\exp\bigg[\frac{(l-\mu)^2}{2\sigma^2}\bigg]
$$

@@

Growth in subsequent age classes was modelled using a length transition matrix [@punt_estimating_1997; @hall_modelling_2000], $G = g_{j,k}$ that represents the probability that a fish in length class, $j$, will grow into length class, $k$, over a specified time interval $$ G = \begin{bmatrix}
 g_{1,1}&0& 0 &0\\ 
 g_{2,1}& g_{2,2} & 0  &0 \\ 
 \cdots&\cdots  & \ddots  & 0\\ 
 g_{n,1}& g_{n,2} & \cdots & g_{n,n}
\end{bmatrix}$$

where the elements of the matrix followed the general form of Punt *et al* [-@punt_estimating_1997] and are given by$$ g_{k,j} = \left\{
    \begin{array}{ll}
        \int_{{\infty}^-}^{\mathrm{l_{k}^+}}f[\phi(l,j)]dl & \mbox{if }k=1 \\
        \int_{l_{k}^-}^{\mathrm{l_{k}^+}}f[\phi(l,j)]dl &\mbox{if }k=1<k<n \\
\int_{l_{k}^-}^{\mathrm{\infty^+}}f[\phi(l,j)]dl &\mbox{if }k=n \\
    \end{array}
\right.
$$

@@

Selectivity in the model was assumed to follow a gamma function based on gillnet selectivity experiments [@kirkwood_gill_1986; @punt_stock_1998], where the selectivity of an individual in length class, $l$ was given by $$S(l) = \left [ \frac{l}{\theta_1 \theta_2} \right ]^{\theta_1 }exp(\theta_1-\frac{l}{\theta_2}) $$ where $\theta_1$ and $\theta_2$ are estimated parameters [@kirkwood_gill_1986]. Two selectivity scenarios were simulated corresponding to low and high selectivity of the maternal component of the population. For consistency between species, gillnet mesh sizes were chosen that would result in peak relative selectivity occurring at the lengths equal to 25% and 75% of the population in maternal condition (Table 1).

@@

$$ N_{l,a} = \left\{    \begin{array}{ll}        1 & \mbox{if }a=0 \\        N_{l,a-1}\exp(-Z_{l}) &\mbox{if }a=1<a<A_{Max} \\        N_{l,a-1}\exp(-Z_{l})/(1-\exp(-Z_{l})) &\mbox{if }a=A_{Max} \\    \end{array}\right.$$

where $A_{Max}$ is the maximum age (Table 1). The estimated numbers of fish caught in length class, l, at age, a, was calculated using the Baranov catch equation as

@@

Next, maturity-at-length data, $\psi(l)$, were randomly generated by simulating from a Bernoulli distribution, $\Psi(l) \sim B(P)$, where $P$ was given by the 2PLF: $$P =  \left ( 1 + e^{-ln(19)\left ( \frac{l - L_{50}}{L_{95} - L_{50}} \right )}  \right )^{-1}$$ This processe was then repeated to generate maternity-at-length data, $\Psi'(l)$, using the 3PLF: $$\Psi' \sim B(P)$$ $$P =  P_{Max}\left ( 1 + e^{-ln(19)\left ( \frac{l - L'_{50}}{L'_{95} - L'_{50}} \right )}  \right )^{-1}$$ Finally, 300 iterations were run for each of the 144 unique variables combinations.

@@

Two approaches to estimating maternity parameters were compared: the 3PLF itself (3PLF-estimated) and the 3PLF function with a fixed asymptote (3PLF-fixed). Additionally, two approaches using a 2PLF were also examined, using maternity-at-length data (2PLF-maternity) and maturity-at-length data (2PLF maturity). The 3PLF-estimated method was undertaken to validate the utility of this model, which has so far been used in only a single study [@colonello_reproductive_2016]. The 3PLF-fixed is the method described by Walker [-@walker_reproduction_2005] and most commonly used in practice. The 2PLF-maternity method was used to examine what effect simply ignoring the upper asmyptote had (i.e. $P_{Max}$ fixed at 1). The 2PLF-maturity method is commonly used to approximate a maternity curve, and tacitly assumed to be similar. The resulting maturity curve can then be weighted by the proportion of gravid females to calculate annual reproductive output (although in practice annual fecundity is more often modified, for example halved for a species suspected to reproduce biennially [@sedar_sedar_2023]).

Only the 3PLF-estimated method involves statistically estimating $P_{Max}$, which must be subjectively chosen in the case of the 3PLF-fixed and the 2PLF-maturity. $P_{Max}$ is ideally chosen based on detailed study of the ovarian and uterine cycles [@walker_reproduction_2005], or alternatively based on the proportion of mature females observed to be in maternal condition during sampling [@baremore_reproduction_2012; @harry_age_2013; @trinnie_reproductive_2016]. For the purposes of the simulation, the timing and duration of the uterine and ovarian cycles were assumed to be unknown. The following procedure was used to 'guess' the fixed value of $P_{Max}$ in each simulation. $P_{Max}$ was chosen as the proportion of females in maternal condition above a the length at which $99\%$ of females were mature, $L > \Psi(L_{99\%})$. In some cases this procedure failed due to there being no maternal females in the simulated data set meeting this criteria. If this occurred, $L_{95}$ was used, followed by $L_{50}$, and finally the proportion of all mature females in the sample that were in maternal condition. Statistical estimation of all non-fixed parameters in each of the methods was undertaken using maximum likelihood. Nonparametric bootstrapping was used to calculate approximate 50% confidence intervals for the best-fit parameters from 250 resampled data sets.

@@

An empirical case study was also undertaken to show how maternity functions can be used in practice. Data from two studies of the reproductive biology of the sandbar shark, *Carcharhinus plumbeus*, in the Gulf of Mexico and western North Atlantic Ocean were combined and reanalysed using both 3PLF methods. The sandbar shark is a relatively long-lived ($A_{Max}$ = 33 years) and slow growing species ($K = 0.12 \text{yr}^{-1}$) with a protracted reproductive cycle lasting longer than a year [@springer_natural_1960]. The Gulf of Mexico and western North Atlantic Ocean population has been considered overfished since the late 1970s [@sminkey_age_1995] and during the 2000s detailed studies of sandbar shark reproductive biology and growth were undertaken for stock assessment purposes [@baremore_reproduction_2012; @piercy_histological_2016]. In their study, Baremore and Hale [-@baremore_reproduction_2012] estimated maternity parameters using the 2PLF-maternity method, weighting the final curve by 0.37 based on the proportion of pregnant females observed, most closely aligning with a triennial reproductive cycle. Piercy *et al* [-@piercy_histological_2016] did not estimate maternity parameters but also suggested that the average reproductive cycle was likely to be longer than two years based on examination of ovarian follicles. To examine relative support for either a biennial or triennial reproductive cycle, 3PLF-fixed models were fit with $P_{Max}$ estimated and fixed at 0.5 and 0.33, respectively.

@@

Parameter estimation was generally straightforward; with sample sizes $> 50$ a success rate of $\geq 95\%$ was achieved in all scenarios (Figure S1, Table S1). Overall, convergence and parameter boundary failures were most common with the 2PLF-maternity method which essentially involved fitting data to an under-parameterized model where $P_{Max}$ was always fixed at 1. In addition to convergence failures, 12 simulations also generated insufficient maternal data (zero or one maternal females in the simulated data), and were discarded.

@@

For the 3PLF methods, parameter estimates tended to be biased high at lower sample sizes (Figure S2, Figure S3). In some scenarios, particularly with sample sizes $\leq$ 250, fixing the asymptote was effective at reducing bias, although for larger sample sizes neither method was clearly preferable (Figure S2, Figure S3). The accuracy of parameter estimates varied considerably across different combinations of variables used in the simulation and among parameters (Figure S4, Figure S5). Again, in some scenarios with smaller sample sizes $\leq$ 250, fixing the asymptote improved precision with the 3PLF methods, although at larger sample sizes better precision was achieved by estimating $P_{Max}$ (Figure 3). Overall, $P_{Max}$ was comparatively more challenging to estimate accurately than $L'_{50}$ (Figure S4, Figure S5).

@@

data %>% 
  filter(par_name %in% "c") %>% 
  filter(method_long %in% c("3PLF - estimated", "3PLF - fixed")) %>% 
  filter(mesh_name %in% "high") %>% 
  ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(as.factor(Nsamples), abs_error, col = repro_freq, group = repro_freq)) +
  geom_line(aes(as.factor(Nsamples), abs_error, col = repro_freq, group = repro_freq)) + 
  facet_grid(species ~ method_long) + 
  scale_color_brewer(palette = "Spectral") + 
  guides(colour = guide_legend(bquote(atop(italic(P)[Max],("Reproductive frequency"))))) + 
  theme_bw() + 
  labs(x = "Sample size", y = "Mean % absolute relative error") +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = "center")
```

While the 3PLF-fixed method was in some cases able to reduce bias at lower sample sizes, a trade off in the use of this method was poorer interval coverage (Figure 4, Figure S6). For most simulations, interval coverage for $L'_{50}$ was well below the expected level of 50%, irrespective of sample size. In contrast, for the 3PLF-estimated method interval coverage oscillated around 50% for both $L'_{50}$ and $P_{Max}$.

```{r fig.width = 7, fig.height= 5.5, fig.cap = "Interval coverage for $\\hat{L'_{50}}$ for 3PLF methods (high selectivity scenarios). Figure shows the percentage of simulations where the true parameter value fell within the 50% bootstrap confidence interval. "}
data %>% 
  filter(par_name %in% "m50", method_long %in% c("3PLF - estimated", "3PLF - fixed")) %>% 
  filter(mesh_name %in% "high") %>% 
  ggplot() +
  geom_hline(yintercept = 50, linetype = "dashed") +
  geom_point(aes(as.factor(Nsamples), int_coverage, col = repro_freq, group = repro_freq)) +
  geom_line(aes(as.factor(Nsamples), int_coverage, col = repro_freq, group = repro_freq)) + 
  facet_grid(species ~ method_long, scales = "free_y") + 
  scale_color_brewer(palette = "Spectral") + 
  guides(colour = guide_legend(bquote(atop(italic(P)[Max],("Reproductive frequency"))))) + 
  theme_bw() + 
  labs(x = "Sample size", y = "% interval coverage") +
  theme(legend.position = "bottom", legend.direction = "horizontal", legend.justification = "center")
```

Differences in life history and gear selectivity played an important role in the ability to obtain precise and unbiased maternity parameters. Certain combinations of variables in the simulated data resulted in captures of few females in maternal or immature condition making parameter estimation difficult (Figure S8, Figure S9). These effects were not necessarily consistent between species and seemed to reflect differences in the underlying population length structure. For example, in low selectivity scenarios for gummy sharks \~60% of individuals were immature  compared to \~50% in the corresponding scenarios for school sharks (Figure S9). As $P_{Max}$ decreased the proportion of maternal females in the analysis also decreased.

Overall, the 2PLF-maternity method performed the worst. Ignoring $P_{Max}$ introduced a fixed bias in this parameter that increased in magnitude as reproductive periodicity increased (Figure S3). This manifested in an overestimation of $L'_{50}$ that was exacerbated by gear selectivity effects (Figure S2). In contrast, using a maturity function to approximate the maternity function (2PLF-maturity method) resulted in relatively good performance. $L_{50}$ could usually be estimated with a higher precision than $L'_{50}$. For the gummy shark, where $L_{50}$ and $L'_{50}$ were relatively similar, using the 2PLF-maturity method led to a constant underestimate of $L'_{50}$ of approximately 2% and exceeding the performance of the 3PLF-methods at most sample sizes.

The effect of the different methods in ultimately calculating $R_0$ varied considerably and was also influenced by life history and selectivity characteristics. For school sharks, estimating $P_{Max}$, was the most effective way to minimize bias in most scenarios for the school shark (Figure 5, Figure S11). In contrast, using a fixed $P_{Max}$ or substituting maturity parameters often led to better performance for the gummy shark (Figure 5, Figure S11).

```{r fig.width = 9, fig.height= 3.5, fig.cap = "Performance of alternative maternity functions in minimising bias in calculations of $R_0$. The preferred method was that which minimised bias, $|\\text{relative error}|$ across 300 simulated datasets. Note 2PLF-maternity (Annual) scenarios were excluded for this comparison."}
